---
title: 'Assignment1_Ramzi_Chariag'
author: "Ramzi Chairag"
date: "Feburary 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, exercise1, eval=FALSE}
3         # -> 3
7 %% 3    # -> 1
False     # -> Error: object 'False' not found
'pi'      # -> "pi"
FALSE     # -> FALSE
```

**Exercise:** Look up what 'snake case' is and add your answer here.
**Answer** Snake case is when_you_name_variables_like_this

**Exercise:** What types of names are allowed in R? Look it up and write your answer here as you understand it. Then provide 3 examples of things that are not valid names in R for different reasons.

**Answer** Object names must start with a letter, and can only contain letters, numbers, _ and .
invalid examples of variable names: v@riable, $tormtrooper, chi`ganshina 

#### Everything is an object

There are built-in objects and objects you import or create. Most importantly, you assign values to objects you create with the '<-' operator: a '<' ('less than') followed by a '-' ('minus' or 'dash').

```{r}
ramzis_new_object <- 666
ramzis_new_object
```

One important function to create objects is the $c()$ function, which combines several items into one object:

```{r}
ramzis_new_combined_object <- c(6,6,6)
ramzis_new_combined_object
```
**Exercise:** What happens if you call the function $mean$ without any arguments, i.e. $mean()$?

```{r mean, eval=FALSE}
mean()
```
**Exercise:** What do 'class()' and 'str()' do? Use 'help' (or '?'). Don't spend too much time reading the docs. Which description do you find more helpful?
```{r}
?class()
?str()
```

## How Ggplot Works

```{r, load, warning=FALSE}
library(tidyverse)
#install.packages("tidyverse")
```

**Exercise:** What other types of plots are there? Try to find several more `geom_` functions.

## Mappings Link Data to Things You See

```{r}
library(gapminder)
library(ggplot2)
gapminder

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()+ geom_smooth()
p + geom_violin()
ggplot(data = gapminder, mapping = aes(x = lifeExp, fill = continent, colour = continent)) + geom_density(alpha = 0.1)
```

```{r include=FALSE}
str(p)
str(p + geom_point())
```

**Exercise:** Make sure that your knitted version doesn't include all the output from the `str(...)` commands, it's too tedious. (Hint: click on the 'gear' symbol in the code block above and then on the link to 'Chunk Options' to get help with chunk options.)

Finally, we add a *layer*. This says how some data gets turned into concrete visual aspects.

```{r}
p + geom_point()
p + geom_smooth()
```


```{r, scale_coordinates}
p + geom_point() + geom_smooth(method = "lm") + scale_x_log10()
```

**Exercise:** Describe what the `scale_x_log10()` does. Why is it a more evenly distributed cloud of points now? (2-3 sentences.)

**Answer** Using a log scale will show logarithmic functions as linear ones on the plot i.e. the x axis is stretched the closer x is to zero and it is more and more compact as the values increase

Nice! The x-axis now has scientific notation, let's change that.

```{r, scales}
library(scales)
p + geom_point() + 
  geom_smooth(method = "lm", color = 'orange') + 
  scale_x_log10(labels = scales::dollar) 

```

**Exercise:** What does the `dollar()` call do?

**Answer** It changes the labels from scientific notation to dollar amounts

```{r, dollar_answer}
?dollar()
```

**Exercise:** How can you find other ways of relabeling the scales when using `scale_x_log10()`?

**Answer** We can use other pre-defined calls like `comma()`, or we can use a vector of labels of our choosing provided that it has the same length as the breaks vector. We can also use a function that takes as input the breaks vector and outputs the labels vector.

```{r,Other_labels}
p + geom_point() + 
  geom_smooth(method = "lm") + 
  scale_x_log10(labels = scales::comma) 

p + geom_point() + 
  geom_smooth(method = "lm") + 
  scale_x_log10(breaks = c(1000, 10000, 100000),labels = c('$1k', '$10k', '$100k')) 
```

### The ggplot Recipe

1. Tell the `ggplot()` function what our data is.
2. Tell `ggplot()` *what* variables we want to map to features of plots. For convenience we will put the results of the first two steps in an object called `p`.
3. Tell `ggplot` *how* to display relationships in our data.
4. Layer on geoms as needed, by adding them on the `p` object one at a time.
5. Use some additional functions to adjust scales, labels, tickmarks, titles. 
  - The `scale_`, `labs()`, and `guides()` functions

### Mapping Aesthetics vs Setting them

```{r, mapping_vs_setting}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = 'yellow'))

p + geom_point() + scale_x_log10()
```

This is interesting (or annoying): the points are not yellow. How can we tell ggplot to draw yellow points?

```{r, yellow_points}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = year))
p + geom_point() + scale_x_log10()

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(color = 'yellow') + scale_x_log10()
```

**Exercise:** Based on the discussion in Chapter 3 of *Data Visualization* (read it), describe in your words what is going on. 
**Answer** Mapping aesthetics in ggplot have do with actual information from the data. We can use it to set the colour according to another variable like year or continent, but we cannot use it to choose a random colour or our choice for the sake of looking nice. In order to impose a colour of our choosing on the graph, we have to do it at a lower level which is the `geom_point()` level.   

One way to avoid such mistakes is to read arguments inside `aes(<property> = <variable>)`as *the property <property> in the graph is determined by the data in <variable>*. 

**Exercise:** Write the above sentence for the original call `aes(x = gdpPercap, y = lifeExp, color = 'yellow')`.

Aesthetics convey information about a variable in the dataset, whereas setting the color of all points to yellow conveys no information about the dataset - it changes the appearance of the plot in a way that is independent of the underlying data. 

Remember: `color = 'yellow'` and `aes(color = 'yellow')` are very different, and the second makes usually no sense, as `'yellow'` is treated as *data* that has the same value always -- namely the value 'yellow'.

```{r, exercise_args_for_smooth}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = year))
p + geom_point() + geom_smooth(color = "orange", se = FALSE, size = 1, method = "lm", weight="year") + scale_x_log10(breaks = c(1000, 10000, 100000),labels = c('$1k', '$10k', '$100k'))
```

**Exercise:** Write down what all those arguments in `geom_smooth(...)` do.
**Answer** method, as the name suggests, has to do with the method by which the smoothing is done. we can choose among a variety of methods that include a kernel density estimator and a linear line. We can actually add weights as well.

```{r, gapminder_with_labels}
p + geom_point(alpha = 0.3) + 
  geom_smooth(method = "gam") + 
  scale_x_log10(labels = scales::dollar) +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
       title = "Economic Growth and Life Expectancy",
       subtitle = "Data Points are country-years",
       caption = "Source: Gapminder")
```

Coloring by continent:

```{r, gapminder_color_by_continent}
library(scales)
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = continent, fill = continent))
p + geom_point()
p + geom_point() + scale_x_log10(labels = dollar)
p + geom_point() + scale_x_log10(labels = dollar) + geom_smooth()
```

**Exercise:** What does `fill = continent` do? What do you think about the match of colors between lines and error bands?
**Answer** `fill = continent` colours the standard error areas and the areas underneath the curve in case we plot a density function.

```{r, gapminder_color_by_continent_single_line}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + geom_smooth() + scale_x_log10()
```

**Exercise:** Notice how the above code leads to a single smooth line, not one per continent. Why?
**Answer** It is because `aes()` at the ggplot level only has x and y as arguments, and the `color = continent` is implemented only at on the scatter plot points. i.e. `geom_smooth()` does not know that the points are coloured by continent. 

**Exercise:** What is bad about the following example, assuming the graph is the one we want? This is why you should set aesthetics at the top level rather than at the individual geometry level if that's your intent.
**Answer** This graph with only one line does not allow us to compare continents properly. having a different line for each continent conveys more information. The graph where we set aesthetics separately for `geom_point()` and `geom_smooth()`

```{r, many_continents}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + 
  geom_smooth(mapping = aes(color = continent, fill = continent), method = "gam") + 
  scale_x_log10()
```

## Additional (not Optional) Exercises 

**Exercise:** Find ways to save the figures that you made so that you can use them elsewhere too. Create a new folder to save only images. Use the command for saving to save the picture for the last image in your new folder, after you have updated the axes, title, subtitle, and caption of the image. Post your solution on Slack and use it to include the final image above with a caption saying "Saved by <your-name>" inside your Slack message (see https://slack.com/help/articles/4403914924435-Add-descriptions-to-images). (Hint: `??save`)

```{r, Ex1}
#jpeg(file="Images/Life_expectancy.jpeg")
ggplot(data = gapminder, mapping = aes(x = lifeExp, fill = continent, colour = continent)) + geom_density(alpha = 0.1) +   labs(x = "Life Expectancy in Years", y = "Density",
       title = "Life Expectancy Distribution by Continent",
       subtitle = "In Africa, we are here for a good time, not a long time",
       caption = "Source: Gapminder")
ggsave(filename = "Images/Life_expectancy.jpeg")

#jpeg(file="Images/Life_expectancy_scatter.jpeg")
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(aes(colour = continent), alpha = 0.4) + 
  geom_smooth(method = "gam") + 
  scale_x_log10(labels = scales::dollar) +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
       title = "Economic Growth and Life Expectancy",
       subtitle = "Data Points are country-years",
       caption = "Source: Gapminder")
ggsave(filename = "Images/Life_expectancy_scatter.jpeg")
```

**Exercise:** Read section 3.8 "Where to go next" from DV. Based on those ideas, experiment and create two different graphs with the gapminder data. Describe each briefly in one sentence.
```{r, Ex2}
ggplot(data = gapminder, mapping = aes(x = lifeExp, fill = continent, colour = continent)) + geom_density(alpha = 0.1) +   labs(x = "Life Expectancy in Years", y = "Density",
       title = "Life Expectancy Distribution by Continent",
       subtitle = "In Africa, we are here for a good time, not a long time",
       caption = "Source: Gapminder")

p <- ggplot(data = gapminder,
            mapping = aes(x = year, y = lifeExp, color = continent))
p + #geom_point(alpha = 0.4) + 
  scale_x_log10() +
  geom_smooth(method = "lm") +
  labs(x = "Year", y = "Life Expectancy in Years",
       title = "Economic Growth and Life Expectancy",
       subtitle = "Data Points are country-years",
       caption = "Source: Gapminder")
```
Description:
the first graph is the pdf for each continent, which allows us to visibly compare the expected values and standard deviations of life expectancy between continents. The second graph is is a linear regression line of life expectancy on time which allows us to compare average level of change over time in life expectancy between continents e.g. life expectancy increased faster in Asia than it did in other continents.

**Exercise:** Read section 1.6 of [R for Data Science](https://r4ds.had.co.nz/introduction.html) on *Getting help and learning more*. Go back to an error from your previous assignment -- or pick a new one -- and post a reproducible error as described in that section on the discourse forum.

**Answer** The following will give an error that 'Breaks and labels are different lengths' the reason is that the default breaks vector is not of size 3. I ran into this error earlier and I sent this code to a friend asking why it does not work, and then I was able to fix it by setting the breaks manually.

```{r, exercise_labels,eval=FALSE}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = year))
p + geom_point() + geom_smooth(color = "orange", se = FALSE, size = 1, method = "lm", weight="year") + scale_x_log10(labels = c('$1k', '$10k', '$100k'))
```

**Exercise:** Do exercise 3.2.4 from [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html#first-steps). Include your code in chunks, describe the output and code (where necessary) in the surrounding text.

```{r, Exercise 3.2.4}
mpg
?mpg
p <- ggplot(data = mpg,
            mapping = aes(x = class, y = drv))
p + geom_point()
#I added this graph out of curiosity, ignore it.
p <- ggplot(data = mpg,
            mapping = aes(x = class, y = hwy, color = class, fill = class))
p + geom_violin(alpha = 0.2) +
geom_violin(alpha = 1, data = mpg,mapping = aes(x = class, y = cty) )

```
mpg is a dataset of Fuel economy data from 1999 to 2008 for 38 popular models of cars. It has 11 variables (columns) and 234 observations (rows).
the variable drv is the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4 wheel drive
the scatter plot is almost useless, because both variables are categorical. The only information we can get is something like "there are no forward wh or 4 wh 2 seater cars" i.e. if there is a point, then there is at least one observation with that combination of wheel drive and class. 

**Exercise:** Go through Exercises in 3.3.1. If an exercise does not make sense immediately (such as you don't know what a categorical variable is), replace the question by a question that addresses that point (in the case of the categorical variable "What are categorical and continuous variables and how are they different in R?"). Write it down, try to answer that question, and ignore the original question. That way you don't end up spending too much time on this one exercise.

```{r, Exercise 3.3.1}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
 

ggplot(data = mpg, mapping = aes(color = cty)) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size = cty, shape = fl, stroke = 1))

ggplot(data = mpg, mapping = aes(color = class)) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ<5, size = cty, shape = fl, stroke = 1))
 

```
The points are not blue because we passed the colour aesthetic in mapping while color doesn’t convey information about a variable, but only changes the appearance of the plot. It make it work, it should go outside of aes().
Categorical variables: The variables that are qualitative i.e. they convey information about a set of different categories, and in mpg they are Manufacturer, model, trans, drv, fl and class.

Size and Color work best with continuous variables while shape works best with categorical variables, provided that there are 6 or less categories.

When you map the same variable to multiple aesthetics, the one used at a lower lever will dominate i.e. we get the one from `geom_point` not the one specified in `ggplot` However, the variable specified at the higher level will still appear on the graph key. In this example the actual colour is based on class, but above the list of different possible classes and their colours in the key, it says "cty" which is the variable specified in ggplot.

What does stroke aesthetic do? 
Stroke controls the thickness of the lines that make up the symbols in the graph.

`aes(colour = displ < 5)` will split the sample into two. one part where the condition is true, and another where the condition is false, and those two parts would be coloured accordingly.

**Exercise:** Read the (very short) [Chapter 4 of R for Data Science](https://r4ds.had.co.nz/workflow-basics.html) and try exercise 1 in section 4.4.
```{r, Exercise 4.4}
library(tidyverse)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)
```
Keyboard shortcuts can be pretty handy. I try not to use the mouse at all most of the time. 

**Bonus Exercise:** Why did I load the `scales` library twice via `library(scales)` to knit? 

## Assignment 1

1. Do the exercises in these and the previous lecture notes.
1. Knit lectures 2, making sure to get rid of those `eval=FALSE` that are just there because I didn't complete the code
1. Submit the pdf on the server